A Signal Averaging Procedure based on Gambini's Algorithm

### 2019-08-26 gsync.tex
Modifico el paper para empezar a documetnar toda la informacion para el algoritmo de sincronizacion de seniales

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
\usepackage[pdftex]{graphicx}
\usepackage{subfig}

  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Histogram of Gradient Orientations of Signal Plots applied to P300 Detection}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Rodrigo~Ramele,~\IEEEmembership{Member,~IEEE,}
        Ana~Julia~Villar,
        and~Juan~Miguel~Santos% <-this % stops a space
\thanks{R. Ramele, A.J.Villar and J.M.Santos are with the Department
of  Computer Engineering, Instituto Tecn{\'o}gico de Buenos Aires (ITBA), Ciudad de Buenos Aires,  Argentina e-mail: rramele@itba.edu.ar}% <-this % stops a space
\thanks{Manuscript received ----; revised ---.}}

@@

% The paper headers
\markboth{IEEE Transactions on Biomedical Engineering}%
{Ramele \MakeLowercase{\textit{et al.}}: Histogram of Gradient Orientations of Signal Plots}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
250 words objective, methods, results, conclusion and significance
The analysis of Electroencephalographic (EEG) signals is of ulterior importance to aid in the diagnosis of mental disease and to increase our understanding of the brain.  Traditionally, clinical EEG has been analyzed in terms of temporal waveforms, looking at rhythms in spontaneous activity, subjectively identifying troughs and peaks in Event-Related Potentials (ERP), or by studying graphoelements in pathological sleep stages.
Additionally, the discipline of Brain Computer Interfaces requires new methods to decode patterns from non-invasive EEG signals.  This field is developing alternative communication pathways to transmit volitional information from the Central Nervous System. The technology could potentially enhance the quality of life of patients affected by neurodegenerative disorders and other mental illness. This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from images of signal plots.  These features are constructed based on the calculation of histograms of oriented gradients from pixels around the signal plot. It aims to provide a new objective framework to analyze, characterize and classify EEG signal waveforms. The feasibility of the method is outlined by detecting the P300, an ERP elicited by the oddball paradigm of rare events, and implementing an offline P300-based BCI Speller.  The validity of the proposal is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset of healthy subjects.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
electroencephalography, histogram of gradient orientations, brain-computer interfaces, P300, SIFT, amyotrophic lateral sclerosis, naive-bayes near neighbours, waveforms
\end{IEEEkeywords}

@@

A few works have explored the idea of exploiting the signal waveform to analyze the EEG signal.  In~\cite{Alvarado-Gonzalez2016} an approach based on Slope Horizontal Chain Code is presented, whereas in~\cite{Yamaguchi2009} a similar procedure was implemented based on Mathematical Morphological Analysis.  The seminal work of Bandt-Pompe Permutation Entropy~\cite{Berger2017} also explores succinctly this idea as a basis to establish the time series ordinal patterns.  In the article~\cite{Ramele2016},  the authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves  and Motor Imagery Rolandic Central $\mu$ Rhythms using the Histogram of Gradient Orientations of signal plots.  Inspired in that work, we propose a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  
The proposed approach is based on the waveform analysis of the shape of the EEG signal.  The signal is drawn on a bidimensional image plot, vector gradients of pixels around the plot are obtained, and with them, the histogram of their orientations is calculated.  This histogram is a direct representation of the waveform of the signal. The method is built by mimicking what regularly electroencephalographers have been performing for almost a century as it is described in~\cite{Hartman2005}: visually inspecting raw signal plots.

@@

\section{Materials and Methods}

The P300~\cite{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a Signal to Noise Ratio (SNR) of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\cite{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\cite{Farwell1988}. This matrix is composed of 6 rows and 6 columns of numbers and letters. The subject can focus on one character of the matrix. Figure~\ref{fig:p300matrix} shows an example of the Speller Matrix used in the OpenVibe open source software~\cite{Renard2010}, where the flashes of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or a column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{openvibep300matrix.png}
\caption{Example of the $6 \times 6$ Speller Matrix used in the study obtained from the OpenVibe software.  Rows and columns flash in random permutations.}
\label{fig:p300matrix}
\end{figure}

@@

\begin{figure*}[!t]
\centering
\subfloat[Classification]{\includegraphics[width=15cm]{classificationgraph.pdf}}
\caption{For each column and row, an averaged, standardized and scaled signal $\tilde{x}^l(n,c)$ is obtained from the segments $S_i^l$  corresponding to the $k_a$ intensification sequences with $ 1 \leq i \leq k_a $ and location $l$ varying between $1$ and $12$. From the averaged signal, the image $I^{(l,c)}$ of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure*}

@@

\begin{itemize}
\item \textbf{Signal Enhancement}: This stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The pipeline starts by applying a notch filter to the raw digital signal, a $4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \cite{Krusienski2006}.
\item \textbf{Artifact Removal}: For every complete sequence of $12$ intensifications of $6$ rows and $6$ columns, a basic artifact elimination procedure is implemented by removing the entire sequence when any signal deviates  $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ intensifications of one intensification sequence,  a segment $S_{i}^l$  of a window of $t_{max} $ seconds of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification $l$ and to the intensification sequence $i$. As intensifications are permuted in a random order, the segments are rearranged corresponding to row flickering, labeled 1-6, whereas those corresponding to column flickering are labeled 7-12.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}: \label{Average}  The P300 ERP is deeply buried under basal EEG so the standard approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out~\cite{Liang2008}.

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the segments $S_1^l(n,c),\dots,S_{k_a}^l(n,c)$, with $1 \leq l \leq 12$, of the EEG signal where the variables $1 \leq n \leq n_{max}$ and $1 \leq c \leq C$  correspond to sample points and channel, respectively. The parameter $C$ is the number of available EEG channels whereas $n_{max}=F_s \  t_{max}$ is the segment length and $F_s$ is the sampling frequency.  The parameter $k_a$ is the number of repetitions of intensifications and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x^l(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i^l(n,c) 
\label{averaging}
\end{equation}  
for $1 \leq n \leq n_{max}$ and for the channels $1 \leq c \leq C$.  This provide an averaged signal $x^l(n,c)$ for the twelve locations $ 1 \leq l \leq 12$.
\end{enumerate}
\end{itemize}

@@

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related to the image scale. In addition, $ x^l(n,c) $ is the point-to-point averaged multichannel EEG signal for the sample point $n$ and for channel $c$. Lastly, $$\bar{x}^l(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x^l(n,c)$$ and $$ \hat{\sigma}^l(c) = \Bigg\{ \frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}} \left[  x^l(n,c)-\bar{x}^l(c) \right] ^2 \Bigg\}  ^{\frac{1}{2}}$$ are the mean and estimated standard deviation of $x^l(n,c), 1 \leq n \leq n_{max}$, for each channel $c$.

@@

\begin{equation}
I^{(l,c)}(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \   n \;  \text{and} \;  z_2 = \tilde{x}^l(n,c) + z^l(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

@@

%\begin{equation}
%z^l(c) = \left \lfloor{ \frac{\max_{n} \tilde{x}^l(n,c)  - \min_{n} \tilde{x}^l(n,c) }{2} }\right \rfloor -   //
%\left \lfloor{ \frac{\max_{n} \tilde{x}^l(n,c)  + \min_{n} \tilde{x}^l(n,c)}{ 2} }\right \rfloor
%\label{eq:zerolevel}
%\end{equation}

\begin{equation}
z^l(c) = \left \lfloor{ \min_{n} \tilde{x}^l(n,c)  }\right \rfloor
\label{eq:zerolevel}
\end{equation}

\noindent where the minimization and maximization are carried out for $n$ varying between ${1 \leq n\leq n_{max}}$, and $ \lfloor \cdot  \rfloor $ denote the rounding to the smaller nearest integer of the number.  This value represents the length of the positive peak of the signal because the signal is inverted before the plot is constructed.

@@

\noindent  where $\mathbf{p}$ is a pixel from the image $I^{(l,c)}$,  $\theta$ is the angle bin with $ \theta $  belonging to $ \{0, 45, 90, 135, 180, 225, 270, 315\} $,  $ |J(\mathbf{p})| $ is the norm of the gradient vector in the pixel $\mathbf{p}$ and it is computed using finite differences and $\angle J(\mathbf{p}) $ is the angle of the gradient vector.  The scalar $ w_\mathrm{ang}(\cdot) $  and vector $ w_{ij}(\cdot) $ functions are linear interpolations used by~\cite{Lowe2004} and \cite{Vedaldi2010} to provide a weighting contribution to eight adjacent bins.  They are calculated as

@@

%\begin{figure}[h!]
%\centering
%\includegraphics[width=12cm]{gradients.png}\label{samplegradients}
%\caption{ (A) Example of a plot of the signal, a keypoint and the corresponding patch. (B) A scheme of the orientation's histogram computation.  Only the upper-left four blocks are visible.  The first eight orientations of the first block, are labeled from $1$ to $8$ clockwise. The orientation of the second block $ B_{1,2} $ is labeled from $9$ to $16$.  This labeling continues left-to-right, up-down until the eight orientations for all the sixteen blocks are assigned. They form the corresponding descriptor of $128$ coordinates.  The length of each arrow represent the value of the histogram on each direction for each block. (C) Vector field of oriented gradients.  Each pixel is assigned an orientation and magnitude calculated  using finite differences. }
%\label{fig:sampledescriptor}
%\end{figure}

\begin{figure*}[!t]
\centering
\subfloat[Case I]{\includegraphics[width=15cm]{gradients.png}}
\caption{ (A) Example of a plot of the signal, a keypoint and the corresponding patch. (B) A scheme of the orientation's histogram computation.  Only the upper-left four blocks are visible.  The first eight orientations of the first block, are labeled from $1$ to $8$ clockwise. The orientation of the second block $ B_{1,2} $ is labeled from $9$ to $16$.  This labeling continues left-to-right, up-down until the eight orientations for all the sixteen blocks are assigned. They form the corresponding descriptor of $128$ coordinates.  The length of each arrow represent the value of the histogram on each direction for each block. (C) Vector field of oriented gradients.  Each pixel is assigned an orientation and magnitude calculated  using finite differences. }
\label{fig:sampledescriptor}
\end{figure*}

%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}

@@

\section{Experimental Protocol} \label{Protocol}

To verify the validity of the proposed framework and method, the public dataset 008-2014~\cite{Riccio2013} published on the BNCI-Horizon website~\cite{Brunner2014} by  IRCCS Fondazione Santa Lucia, is used. Additionally, an own dataset with the same experimental conditions is generated. Both of them are utilized to perform an offline BCI Simulation to decode the spelled words from the provided signals.

The algorithm is implemented using  VLFeat~\cite{Vedaldi2010} Computer Vision libraries on MATLAB V2014a (Mathworks Inc., Natick, MA, USA). Furthermore, for a sake of reproducibility, the code of the algorithm has been made available at: https://bitbucket.org/itba/hist.

In the following sections the characteristics of the datasets and parameters of the identification algorithm are described.

\subsection{P300 ALS Public Dataset} \label{ALSDataset}

The experimental protocol used to generate this dataset is explained in~\cite{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The Visual P300 detection task designed for this experiment consisted of spelling 7 words of 5 letters each, using the traditional P300 Speller Matrix~\cite{Farwell1988}. The flashing of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 words are used for calibration and the remaining 4 words, for testing with visual feedback.  A trial is every attempt to select a letter from the speller. It is composed of signal segments corresponding to $k_a =10$ repetitions of flashes of 6 rows and $k_a =10$ repetitions of flashes of 6 columns of the matrix, yielding 120 repetitions.  Flashing of a row or a column is performed for 0.125 s, following by a resting period (i.e. inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause is included before resuming with the following letter.

The recorded dataset was sampled at 256 Hz and it consisted of a scalp multichannel EEG signal for electrode channels Fz, Cz, Pz, Oz, P3, P4, PO7 and PO8, identified according to the 10-20 International System,  for each one of the 8 subjects.   The recording device was a research-oriented digital EEG device (g.Mobilab, g.Tec, Austria) and the data acquisition and stimuli delivery were handled by the BCI2000 open source software~\cite{Schalk2004}.

In order to assess and verify the identification of the P300 response, subjects are instructed to perform a copy-spelling task. They have to fix their attention to successive letters for copying a previously determined set of words, in contrast to a free-running operation of the speller where each user decides on its own what letter to choose.

\subsection{P300 for healthy subjects}

We replicate the same experiment on healthy subjects using a wireless digital EEG device (g.Nautilus, g.Tec, Austria).  The experimental conditions are the same as those used for the previous dataset, as detailed in section~\ref{ALSDataset}.  The produced dataset is available in a public online repository~\cite{owndataset}.

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigaci{\'o}n y Doctorado, Instituto Tecnol{\'o}gico de Buenos Aires (ITBA)}.  All healthy subjects have normal or corrected-to-normal vision and no history of neurological disorders. The experiment is performed with 8 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 29.00 years, standard deviation  11.56 years, range 20-56 years.

EEG data is collected in a single recording session. Participants are seated in a comfortable chair, with their vision aligned to a computer screen located one meter in front of them.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\cite{Renard2010}.

Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on the same positions Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is slightly different, and is set to 250 Hz, which is the closest possible to the one used with the other dataset.

%Fz, Cz, P3, Pz, P4, PO7, PO8 and Oz.

%8 gel-based active electrodes (g.LADYbird) + g.LADYbird (GND) + g.GAMMAearclip (REF) C3, Cz, C4, CPz, P3, Pz, P4, POz, GND: AFz, REF: right ear

\subsection{Parameters}

The patch size is $X_P = 12s \times 12s$ pixels, where $s$ is the scale of the local patch and it is an input parameter of the algorithm. The P300 event can have a span of $400$ ms and its amplitude can reach $ 10 \mu V $~\cite{Rao2013}.  Hence it is necessary to utilize a signal segment of size $t_{max} = 1$ second and a size patch $X_P$ that could capture an entire transient event. With this purpose in consideration, the $s$ value election is essential.

%necesitamos definir el valor de s en función de los parámetros de la señal, de modo tal que el parche cubra el evento completo.  
We propose the Equations~\ref{eq:mapping2} and~\ref{eq:mapping1} to compute the scale value in horizontal and vertical directions, respectively. 
\begin{equation}
s_x = \frac{ \gamma \;  \lambda \  F_s}{12}
\label{eq:mapping2}
\end{equation}

\begin{equation}
s_y= \frac{\gamma \; \Delta \mu V}{12} 
\label{eq:mapping1}
\end{equation}

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ F_s $ is the sampling frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $,  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also determines that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ on the vertical direction and $\frac{1}{F_s \ \gamma}=\frac{1}{64}$ seconds on the horizontal direction. The keypoints  $\mathbf{p_k}$  are located at $ (x_{p_k}, y_{p_k} )= ( 0.55 F_s \ \gamma, z^l(c) )= (35,  z^l(c)) $ for the corresponding channel $c$ and location $l$ (see Equation~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:patchgeometry} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction).

\begin{figure}[h!]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole transient event.  The size of the patch is $X_p \times X_p$ pixels. The vertical size consists of $4$ blocks of size $3 s_y$ pixels which is high enough as to contain the signal $\Delta  \mu V $, the peak-to-peak amplitude of the transient event. The horizontal size includes $4$ blocks  of $3 s_x$ and covers the entire duration in seconds of the transient signal event, $ \lambda $.   }
\label{fig:patchgeometry}
\end{figure}

>>>>

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{A Signal Averaging Procedure based on Gambini's Algorithm}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Rodrigo~Ramele,~\IEEEmembership{Member,~IEEE,}
        Juliana~Gambini,
        and~Juan~Miguel~Santos% <-this % stops a space
\thanks{R. Ramele, J.Gambini and J.M.Santos are with the Department
of  Computer Engineering, Instituto Tecn{o'}gico de Buenos Aires (ITBA), Ciudad de Buenos Aires,  Argentina e-mail: rramele@itba.edu.ar}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

@@

% The paper headers
\markboth{IEEE Transactions on Biomedical Engineering}%
{Ramele \MakeLowercase{\textit{et al.}}: A Signal Averaging Procedure based on Gambini's Algorithm}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
EEG
BCI
Signal Averaging
Solutions
What lies behind.

\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
IEEE, IEEEtran, journal, \LaTeX, paper, template.
\end{IEEEkeywords}

@@

A few works have analyzed the problem of syncrhonizing different signals to average them and to enhance their SNR.

@@

@@

@@

\begin{itemize}
\item \textbf{Signal Enhancement}: This stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The pipeline starts by applying a notch filter to the raw digital signal, a $4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \cite{Krusienski2006}.
\item \textbf{Artifact Removal}: For every complete sequence of $12$ intensifications of $6$ rows and $6$ columns, a basic artifact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ intensifications of one intensification sequence,  a segment $S_{i}^l$  of a window of $t_{max} $ seconds of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification $l$ and to the intensification sequence $i$. As intensifications are permuted in a random order, the segments are rearranged corresponding to row flickering, labeled 1-6, whereas those corresponding to column flickering are labeled 7-12.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}: \label{Average}  The P300 ERP is deeply buried under basal EEG so the standard approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out~\cite{Liang2008}.

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the $1 \leq l \leq 12$ segments $S_1^l(n,c),\dots,S_{k_a}^l(n,c)$, of the EEG signal where the variables $1 \leq n \leq n_{max}$ and $1 \leq c \leq C$  correspond to sample points and channel, respectively. The parameter $C$ is the number of available EEG channels whereas $n_{max}=F_s \  t_{max}$ is the segment length and $F_s$ is the sampling frequency.  The parameter $k_a$ is the number of repetitions of intensifications and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x^l(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i^l(n,c) 
\label{averaging}
\end{equation}  
for $1 \leq n \leq n_{max}$ and for the channels $1 \leq c \leq C$.  This provide an averaged signal $x^l(n,c)$ for the twelve locations $ 1 \leq l \leq 12$.
\end{enumerate}
\end{itemize}

@@

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related to the image scale. In addition, $ x^l(n,c) $ is the point-to-point averaged multichannel EEG signal for the sample point $n$ and for channel $c$. Lastly, $$\bar{x}^l(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x^l(n,c)$$ and $$ \hat{\sigma}^l(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x^l(n,c)-\bar{x}^l(c))^2 )^{\frac{1}{2}}$$ are the mean and estimated standard deviation of $x^l(n,c), 1 \leq n \leq n_{max}$, for each channel $c$.

@@

\begin{equation}
I^{(l,c)}(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \  n; \! z_2 = \tilde{x}^l(n,c) + z^l(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

@@

\begin{equation}
z^l(c) = \left \lfloor{ \frac{\max_{n} \tilde{x}^l(n,c)  - \min_{n} \tilde{x}^l(n,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max_{n} \tilde{x}^l(n,c)  + \min_{n} \tilde{x}^l(n,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

\noindent where the minimization and maximization are carried out for $n$ varying between ${1 \leq n\leq n_{max}}$, and $ \lfloor \cdot  \rfloor $ denote the rounding to the smaller nearest integer of the number.

@@

\noindent  where $\mathbf{p}$ is a pixel from the image $I^{(l,c)}$,  $\theta$ is the angle bin with $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $,  $ |J(\mathbf{p})| $ is the norm of the gradient vector in the pixel $\mathbf{p}$ and it is computed using finite differences and $\angle J(\mathbf{p}) $ is the angle of the gradient vector.  The scalar $ w_\mathrm{ang}(\cdot) $  and vector $ w_{ij}(\cdot) $ functions are linear interpolations used by~\cite{Lowe2004} and \cite{Vedaldi2010} to provide a weighting contribution to eight adjacent bins.  They are calculated as

@@

@@

\subsection{Subsection Heading Here}
Subsection text here.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

\section{Conclusion}
The conclusion goes here.
